### 2023-05-25 Manuscript/Draft1.tex
small manuscript update

To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences. We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time. In both cases we analyze the ability of stratified random sampling to track population trends through a design-based approach (stratified mean) compared to a model-based approach that allows for the inclusion of covariates (Vector Autoregressive Spatiotemporal Model).

@@

We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model. The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs. VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and can take on the order of hours to complete.

We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass in our Georges Bank population models using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process running models without covariate information as well as including covariates in our model. We considered covariates in the form of dynamic temperature values and/or static habitat values (\(Hab_{J.s}\)) from our population models. When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we ultimately compared the performance of two sets of settings in our VAST models, which can be seen in Table\ref{tab:VASTsettings}.

@@

The top two panels in Figure \ref{fig:PopPct} depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature. While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow exterior strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow exterior strata 17 in the fall, while seemingly none of the population exists in any of the exterior strata (14, 15, 17, 18) in the spring.

Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. When all strata are sampled, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}. However, Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.

Of all scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate. All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value. These 15 fall estimates span all other scenario variations. The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail Flounder, which is that a larger percentage of the population exists in the narrow exterior strata (18 and/or 17). This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.

@@

Table \ref{tab:CodERROR} contains the absolute error between our abundance estimates for Cod and the true model values. Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate. VAST produced higher relative error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates. Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place. Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean. However, we see that adding covariates to VAST in the fall produces higher absolute error values when sampling reduced strata in the constant temperature scenario.

@@

Figure \ref{fig:PopPct} reveals some subtle seasonal differences in the percent of haddock in each strata. In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 24 and 29. This change represents a northward movement between larger centrally located strata. While strata 24 and 29 also increase in the fall under constant temperature, the corresponding decrease is primarily from strata 16 and 13. While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticeable increases in strata 17, 18, and 29. This shift represents movement from the shallower and more centrally located strata towards exterior deeper strata. Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most strata of 29 in the fall demonstrates a climate-driven change in movement preference.

Table \ref{tab:HadERROR} contains the absolute error between our abundance estimates for Haddock and the true model values. We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean. VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases. For Haddock results, adding covariates improves estimates only when all strata are included. That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy. Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor smaller exterior stratum 17.

@@

\emph{Differences Between VAST Settings (mostly notes, not fully formed)}

For YTF, 29/96 \textasciitilde{} 30\% of VAST runs with new settings were better (70\% were worse). 80/96 \textasciitilde{} 83\% of scenarios had a VAST run with a better error than the stratified mean. 15/16 of the times the stratified mean was better VAST was not using covariates. The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop\_IncTemp\_Allstrata\_WCov\_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP\_IncTemP\_allstrata for Had and YT). The overestimate in the fall only seems to be related to the way the population shifts between season. Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata). In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17). Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates

For Haddock (IncPop), 18/32 \textasciitilde{} 57\% of VAST runs better with new settings. 24/32 = 75\% of scenarios had a VAST run with a better error than the stratified mean. 14/16 VAST were better with covariates (the two that were worse were essentially the same) while only 10/16 were better without covariates. Stratified mean tended to perform better when all strata were included in the calculation, while VAST tended to perform better when strata are removed.

For Cod (DecPop), only 11/32 \textasciitilde{} 34\% of VAST runs were better with new settings. 28/32\textasciitilde{} 88\% of scenarios had a VAST run with a better estimate than the stratified mean. All 16/16 = 100\% of VAST runs with a covariate were better than the stratified mean. 12/16 VAST were better without covariates. The 4 runs without covariates that were worse were all extremely large errors and all from increasing temp in the fall (with and without all strata and noise). In the fall the population is moving out of strata 16 (large eastern strata) and 21 (north large) and into a number of strata (22 (tiny north), 24 (large north), 18 (tiny north)). Might be having trouble tracking in to smaller strata again.

@@

{[}{[}we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others? also should note that funding provided by Northeast Fisheries and Climate program at the NEFSC source (I'll dig up the official name of the funding source){]}{]}

>>>>

To test the ability of the bottom trawl survey to track population trends under shifting environmental conditions, we construct spatial models for fish where movement depend on temperature preferences. We consider the impact of climate change by comparing simulations that use a repeating water temperature pattern to those where temperature increases on average over time. In both cases we conduct stratified random sampling on model output and analyze the ability of the samples to track population trends. We compare yearly abundance estimates obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model. The stratified mean is a design-based approach that calculates the arithmetic mean catch per tow and has traditionally been used with stratified random sample designs. VAST is a spatial delta-generalized linear mixed model that estimates both abundance (biomass) and probability of occurrence (presence/absence) (Thorson 2019). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses.

@@

The design-based method we used is the stratified mean, which divides the inhabited domain into \(N\) disjoint strata based on relevant geographic and environmental information such as depth and latitude/longitude. The number of samples taken from each stratum \(S_j\) for \(j=1...N\) is relative to the given area. The stratified mean biomass \(SM_{s,y}\) for a given season \(s\) and year \(y\) can then be calculated by through the weighted average

\[
  \makebox[\linewidth]{$SM_{s,y} = \sum_{i=1}^N W_i \frac{\sum_{k=1}^{S_i}y_{k,i}}{S_i}$},
\]

where \(W_i=\frac{S_i}{\sum_{j=1}^N S_j}\) and \(y_{k,i}\) represents the biomass in sample \(k\) of stratum \(i\) in season \(s\) and year \(y\). These tow-dependent calculations are quick and easy to calculate, especially relative to the model-based VAST estimates.

We compare stratified mean estimates to those derived from the model-based VAST approach. VAST models both biomass \(p_1(i)\) and presence/absence \(p_2(i)\) for each observation \(i\) as linear predictors using a spatial delta-generalized linear mixed model. The equation representing the biomass predictor can be represented by

\[
  \makebox[\linewidth]{$p_1(i)=\text{temporal variation} + \text{spatial variation} + \text{spatio-temporal variation} + \text{vessel effects} + \text{habitat covariates} + \text{catchability covariates}$},
\]

where more specific functional forms of each component are further described in (Thorson 2019). \(p_2(i)\) takes on a similar form as \(p_1(i)\). VAST models require numerous user inputs to determine how the linear predictors will be conditioned and solved. As a result, VAST models take on the order of hours to complete.

We follow the advice given in (Thorson 2019) to build VAST models to estimate biomass in our Georges Bank population models using stratified random samples from our model output. In addition to exploring different link functions and assumed distributions, our VAST model-building process involved testing the impact of including spatial and/or spatio-temporal variation in our models, considering varying number of knots in our mesh, and testing different forms of temporal correlation. We carried out the same model-building process running models without covariate information as well as including covariates in our model. We considered covariates in the form of dynamic temperature values and/or static habitat values (\(Hab_{J.s}\)) from our population models. When using covariates we ultimately decided to consider a best-case scenario by including both temperature and habitat covariates for both linear predictors in order to provide the most information to the model. Knowing the true population values in our models allowed us to calculate the absolute error of each VAST estimate to compare between potential settings. Through this process, and in consultation with the VAST package creator, we compared the performance of two sets of settings in our VAST models, which can be seen in Table\ref{tab:VASTsettings}. VAST Settings B were chosen after reviewing the information provided in the ``Seasonal Model'' section of VAST's online tutorial \url{https://james-thorson-noaa.github.io/docs/tutorials/seasonal-model/}. After running models for each of our scenarios using these settings we shared some of our results with the VAST package creator, James (Jim) Thorson. Jim reviewed results and suggested we also consider the options in Settings A. Both settings use a Tweedie model and turn off the first linear predictor \(p_1\) so that only biomass is being modeled. The difference between the two settings is in the Rho Configuration setting where RhoBeta2 = 0 in Settings B while RhoBeta2 = 3 in Settings A. Intercept for the biomass predictor is a fixed effect (beta2=0) vs constant among years as a fixed effect (beta2=3)

@@

The top two panels in Figure \ref{fig:PopPct} depict the results for Yellowtail Flounder with a repeating temperature gradient on the left (constant temperature) and a temperature gradient that increases over time on the right (increasing temperature). In both temperature scenarios we see the percent of Yellowtail Flounder in strata 13 decrease over the course of the time series in both seasons. The spring population in stratum 19 also decreases in both temperature scenarios as well. The percent of the population increases in stratum 16 in the spring over the duration of both time series, which implies the flow out of strata 13 and 19 in the spring are going into stratum 16. These dynamics occur in both temperature scenarios in weeks 13 and 14 because stratum 16 contains favorable habitat for Yellowtail Flounder that coincides with most of the areas we have designated as the species' spawning ground, which takes place in weeks 9-12. These spring changes are therefore related to the static habitat values in our model rather than the temperature preferences, which is why we seen the same dynamics with constant and increasing temperature. While we observe similar changes in the fall (weeks 37 and 38) in the constant temperature scenario, an increasing average temperature results in a decrease in the population in stratum 16 over time and corresponding increases in strata 17 and 18 (see Figure \ref{fig:PopPct}). These dynamics imply that an increase in temperature results in the more shallow strata 16 becoming less desirable than the deeper and more narrow outer strata 17 and 18. One noticeable seasonal difference in the constant temperature scenario for Yellowtail is how \textasciitilde10\% of the population exists in the narrow outer strata 17 in the fall, while seemingly none of the population exists in any of the strata near the edge of the domain (14, 15, 17, 18) in the spring.

Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata} contain the absolute error between our abundance estimates for Yellowtail Flounder comparing design-based approach (stratified mean) with two settings for a model-based approach (VAST A \& B). In reviewing these Tables we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. The settings used in VAST mattered with Settings B outperforming Settings A in \(67/96\approx 70\%\) of scenarios. When all strata are sampled, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. There are several instances in which VAST failed to provide improved abundance estimates compared to the stratified mean during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}. However, Including covariate information to each VAST model produced estimates with significantly lower error than their stratified mean counterparts.

Of all scenarios without covariates, \(33/48\approx68\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate. All 15 of the covariate-free VAST estimates that resulted in a higher error than the stratified mean were for the fall season and had a common theme of producing abundance estimates that are above the true model value. These 15 fall estimates span all other scenario variations. The implication of this is that our model-based approach without covariates struggles with the primary seasonal difference for Yellowtail Flounder, which is that a larger percentage of the population exists in the narrow strata toward the edge of the domain (18 and/or 17). This theory is further supported by the fact that the absolute error in the increasing temperature scenarios increased dramatically in the fall season, when the combined percentage of the population in the outer strata 17 and 18 increased to over 40\% by the end of the simulation.

@@

Table \ref{tab:CodERROR} contains the absolute error between our abundance estimates for Cod and the true model values. The VAST settings performed comparably to each other with \(18/32\approx57%
\) of estimates performing better with VAST Settings A. Of the abundance estimates without covariates, \(12/16=75\)\% had a VAST fit with a lower relative error compared to the stratified mean estimate. VAST produced higher relative error compared to the stratified mean in scenarios that involved increasing temperature in the fall season without covariates. Similar to the Yellowtail results, this implies the model-based approach had seasonal trouble with populations shifting into smaller strata where fewer samples take place. Providing covariate information in these cases once again helped the model-based approach to provide improved absolute error estimates relative to the stratified mean. However, we see that adding covariates to VAST in the fall produces higher absolute error values when sampling reduced strata in the constant temperature scenario.

@@

Figure \ref{fig:PopPct} reveals some subtle seasonal differences in the percent of haddock in each strata. In the constant temperature scenarios, the spring shows a decrease in strata 19 and 20 that correspond to increases in 16, 24 and 29. This change represents a northward movement between larger centrally located strata. While strata 24 and 29 also increase in the fall under constant temperature, the corresponding decrease is primarily from strata 16 and 13. While similar results can be seen in the spring for the increasing temperature scenario, much more dramatic results exist in the fall under increasing temperature as we see a significant decreases in strata 13, 16, 21, and 22 that leads to the most noticeable increases in strata 17, 18, and 29. This shift represents movement from the shallower and more centrally located strata towards deeper strata located near the edge of the domain. Since stratum 16 contains very favorable habitat including much of the species' spawning ground, the strong shift out of 16 and into the northern most strata of 29 in the fall demonstrates a climate-driven change in movement preference.

Table \ref{tab:HadERROR} contains the absolute error between our abundance estimates for Haddock and the true model values. Absolute error values tended to be lower with VAST Settings B about \(21/32\approx 66\%\) of scenarios. We notice that the model-based VAST produced particularly large errors in spring compared to the stratified mean, with added covariates only improving to the level of the stratified mean. VAST shows improved results in fall relative to the stratified mean with added covariates producing extremely low errors in some cases. For Haddock results, adding covariates improves estimates only when all strata are included. That is, similar to Cod, when sampling a reduced domain adding covariates actually decreases VAST's accuracy. Since this occurs in all scenarios, it seems to again be related to failing to accurately monitor stratum 17 near the edge of the domain.

@@

@@

{[}{[}we should thank Jim (obviously), but also those who helped you with the habitat data (robyn, david Chev.) and others? also should note that funding provided by Northeast Fisheries and Climate program at the Northeast Fisheries Science Center {]}{]} Also thank Joe Langan for helping develop the proposal that funded this work.
